A Handbook for Tenured Track Faculty in Computer Science in the US

### 2024-06-07 demystify.tex
updates

By around March you should hear back from most PhD programs you applied to. If you haven't heard back, reach out via email and ask about the status of your application.
If you receive offers, congratulations!  Now you're at a different game because the schools that have admitted you will try to get you to accept them!  Important factors to consider include the reputation of schools and professors (\autoref{sec:schoolsandprofs}), and funding availability (\autoref{sec:funding}). You will likely have to make your decision (\autoref{sec:accept-postpone-decline}) by around \emph{April 15}, which is the deadline for most schools
(\href{https://cgsnet.org/wp-content/uploads/2024/01/CGS_April15_Resolution_Jan312024.pdf}{why this date?}).

\paragraph{Open House} Most schools have \emph{Open House} or \emph{Visit Day} events, which are a great resource to learn about the school, department, faculty, research, living, etc.
\textbf{Advice}: Even if you can't come in person, you should attend virtually and meet with individual faculty. During the event, you get a chance to learn more about the program, and talk to individual faculty and current students.  Take notes of faculty who make you excited, and count those taking in new students (if they meet you, likely they are considering new students!).  Talk to students about their advisers, the dept, the area, and the funding situation.  Ask about anything you want to determine that they deserve \emph{you}.

@@

\epigraph{Bart, with \$10,000, we’d be millionaires! We could buy all kinds of useful things like…love!’}{\textsc{The Simpsons}}

As mentioned, if you're admitted to a \emph{good} CS PhD program, you should not have to worry about funding!
In the US, the common types of funding for PhDs are \emph{graduate teaching assistant} (GTA or TA), \emph{graduate research assistant} (GRA or RA), and \emph{Fellowship}.
RA is paid by a prof. for you to do their research. TA is paid by the department for you to help with teaching. Finally, fellowship is independent funding that can come from a school, a company, or an organization. \autoref{tab:funding} summarizes the differences.
Note that funding is typically more available for PhD students than
Masters.

@@

Several things about stipends.  First, the amount of stipend \emph{varies} and depends on factors such as location (e.g., a stipend in Washington DC is likely higher than in Lincoln, Nebraska due to higher living costs). Second, an academic year (AY)  year is typically \emph{9-month} in the US, so the stipend is for 9 months (so divide by 9 for each month). Third, you might get paid over the summer (\autoref{sec:summer-funding}) through funding from your professor or fellowship (typically no TA over the summer). Fourth, like most sources of income in the US, you will have to \emph{pay tax} on your stipend.  Finally, private universities might pay more for stipends, but they could have extra activity fees or some other hidden ones (e.g., you may need to pay some fees for each credit hour).

>>>>

By around March you should hear back from most PhD programs that want to admit you.  In many cases you will not hear back from schools that do not want to admit you.
If you receive offers, congratulations!  Now you're at a different game because the schools that have admitted you will try to get you to accept them!  Important factors to consider include the reputation of schools and professors (\autoref{sec:schoolsandprofs}), and funding availability (\autoref{sec:funding}). You will likely have to make your decision (\autoref{sec:accept-postpone-decline}) by around \emph{April 15}, which is the deadline for most schools
(\href{https://cgsnet.org/wp-content/uploads/2024/01/CGS_April15_Resolution_Jan312024.pdf}{why this date?}).

\paragraph{Open House} Most schools have \emph{Open House} or \emph{Visit Day} events, which are a great resource to learn about the school, department, faculty, research, living, etc.

Even if you can't come in person, you should attend virtually and meet with individual faculty. During the event, you get a chance to learn more about the program, and talk to individual faculty and current students.  Take notes of faculty who make you excited, and count those taking in new students (if they meet you, likely they are considering new students!).  Talk to students about their advisers, the dept, the area, and the funding situation.  Ask about anything you want to determine that they deserve \emph{you}.

@@

\epigraph{Bart, with \$10,000, we’d be millionaires! We could buy all kinds of useful things like…love!’}{\textsc{The Simpsons}}
If you're admitted to a \emph{good} CS PhD program, you should not have to worry about funding!
In the US, the common types of funding for PhDs are \emph{graduate teaching assistant} (GTA or TA), \emph{graduate research assistant} (GRA or RA), and \emph{Fellowship}.
RA is paid by a prof. for you to do their research. TA is paid by the dept. for you to help with teaching. Finally, fellowship is independent funding that can come from a school, a company, or an organization. \autoref{tab:funding} summarizes the differences.
Note that funding is typically more available for PhD students than
MS.

@@

Several things about stipends.  First, the amount of stipend \emph{varies} and depends on factors such as location (e.g., a stipend in the DC area is likely higher than in Lincoln, Nebraska due to higher living costs). Second, an academic year (AY)  year is typically \emph{9-month} in the US, so the stipend is for 9 months (so divide by 9 for each month). Third, you might get paid over the summer (\autoref{sec:summer-funding}) through funding from your professor or fellowship (typically no TA over the summer). Fourth, like most sources of income in the US, you will have to \emph{pay tax} on your stipend.  Finally, private universities might pay more for stipends, but they could have extra activity fees or some other hidden ones (e.g., you may need to pay some fees for each credit hour).

### 2024-06-07 faculty.tex
updates

\chapter{Research}
\chapter{Obtaining Funding}
Getting funding is arguably the most important part of getting tenure.  Most new professors struggle with grant writing, which is very different compared to research papers. However, without funding, you can't pay your students, conference registration and travel, or even pay your own summer salary. In short, the lack of funding causes lost of stress and anxiety, and threaten your tenure.  In this chapter, we will discuss the various ways to obtain funding.

@@

\section{Timeline}
Typically you will first mention to your department chair that you are planning to go up for tenure. The department chair then assign you a mentor who will help you with the process. You will apply for tenure in the Fall semester, and the decision will be made in the Spring semester.

My own timeline is as follows:

\begin{itemize}
    \item Fall 2021: Moved to GMU with accelerated tenure clock (meaning I don't have to restart from the beginning as I have already been at UNL since 2016).
    \item (late) Fall 2022: I told my department chair that I plan to apply for tenure in 2023 (which means if things go well then I get tenured in Fall'24).
    \item Spring 2023: I started preparing my tenure package, which includes writing statements and think reference letters.  I also took a semester off because GMU has a pre-tenure study break policy.
    \item Summer 2023: I submitted my tenure package.
    \item Fall 2024: I gave a tenure talk. Department voted in my favor, and then the college voted in my favor, and the decision goes up to the university.
    \item Spring 2024: In May, I got the official notification that I got tenured and promoted to Associate Prof., beginning in Fall 2024.
\end{itemize}

@@

@@

@@

\paragraph{Tenure Visits} Many faculty will visit other universities to give talks and meet people.  This is a good way to get your name out and to get letters.  For example, if you plan to put person X on your list or you think person X will likely be selected by the department, you can ask X to visit their university and give a talk.

>>>>

\chapter{Research}
\chapter{Obtaining Funding}
Getting funding is arguably the most important part of getting tenure.  Most new professors struggle with grant writing, which is very different compared to research papers. However, without funding, you can't pay your students, conference registration and travel, or even your own summer salary. In short, the lack of funding causes lost of stress, uncertainty, and threaten your career.  In this chapter, we will discuss the various ways to obtain funding.

@@

\section{Timeline}
Typically you will first mention to your department chair that you are planning to go up for tenure. The chair might then assign you a mentor who will help you with the process. You will apply for tenure in the Fall semester, the decision will be made in the next Spring semester, and you will get tenure in the Fall semester after that.

My timeline is as follows:

\begin{itemize}
  \item Fall 2021: Moved to GMU with accelerated tenure clock (meaning I don't have to restart from the beginning as I have already been at UNL since 2016).
  \item (late) Fall 2022: I told my department chair that I plan to apply for tenure in 2023 (which means if things go well then I get tenure in Fall'24).
  \item Spring 2023: I started preparing my tenure package, which includes writing statements and think reference letters.  I also took a semester off because GMU has a pre-tenure study break policy.
  \item Summer 2023: I submitted my tenure package.
  \item Fall 2024: I gave a tenure talk. Department voted in my favor, and then the college voted in my favor, and the decision goes up to the university.
  \item Spring 2024: In May, I got the official notification that I got tenure and promoted to Associate Prof., beginning in Fall 2024.
\end{itemize}

@@

GMU (and likely other universities) has a very strict and specific format for the CV (e.g., what to include, how to format, etc).  Fortunately I have been using this format since I joined GMU and also kept my CV up-to-date, so I didn't have to do much. I did have to add a few more sections (e.g., teaching evaluation and number of students in courses taught).  I also have to make sure that the CV is consistent with my statements and other parts of the package (e.g., funding amount must match with what the university has in their system).

I also looked at CVs of other people at GMU who recently got tenure. I also sent my CV to  my mentors and got some feedback.

@@

Here is my \href{https://dynaroars.github.io/people/nguyenthanhvuh/files/pt/statement-annotated.pdf}{tenure application statement}, annotated with comments. At GMU, the statement is limited to 8 pages.  My statement follows the standard structure and talks about my achievements and plans in (i) research and funding, (ii) advising and mentoring, (iii) teaching, and (iv) services. The main unique part of my statement is the first page where I highlight and summarize my achievements (e.g., papers, fundings, students, etc). I get most of these information and numbers from my CV.

While preparing the statement was the most time-consuming part of my tenure package, it didn't take as long (may be just a couple of days).  I get statements from several colleague at GMU who recently got tenure and a couple of people at other universities to get some ideas on how it should be structured.  I was able to reuse lots of text and contents from my job application statements (when I applied to GMU in 2021).  I did not send it for feedback and also did not revise it much.

@@

\paragraph{Tenure Visits} Many faculty will visit other universities to give talks and meet people.  This is a good way to get your name out and to get letters (e.g., from the person that host you).  Physical visits were not possible during the pandemic, but virtual visits are possible and in fact easier. However,  I did not do any tenure visits or give talks in the year leading up to my tenure application. This was because I am not a travel person and also I felt that it would not help much in my case.
